---
title: "endoSignatureR-intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{endoSignatureR-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>"
)
```

```{r setup}
library(endoSignatureR)
```

# Introduction

`endoSignatureR` is a package for discovering small, interpretable gene signatures from bulk RNA-seq data with binary phenotypes in small cohorts. The package provides three distinct workflows:

- **Mode 1 (Rapid Classification)**: Apply pre-trained signature to new samples
- **Mode 2 (Signature Validation)**: Train and validate signatures on labeled data
- **Mode 3 (Visualization & Analysis)**: Explore and visualize data with QC plots

For detailed workflow demonstrations, see the dedicated vignettes:

- `browseVignettes("endoSignatureR")` - List all available vignettes
- Mode 3: `vignette("mode3-visualization-analysis", package = "endoSignatureR")`
- Mode 2: `vignette("mode2-signature-validation", package = "endoSignatureR")` (coming in Phase 2)
- Mode 1: `vignette("mode1-rapid-classification", package = "endoSignatureR")` (coming in Phase 3)
